---
title: "Darwin Core mapping"
subtitle: "For: alien-species-romania"
author:
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# Setup 

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = TRUE)
```

Install required libraries (only if the libraries have not been installed before):

```{r}
installed <- rownames(installed.packages())
required <- c("tidyverse", "tidylog", "magrittr", "here", "janitor", "readxl", "digest", "rgbif")
if (!all(required %in% installed)) {
  install.packages(required[!required %in% installed])
}
```

Load libraries:

```{r message = FALSE}
library(tidyverse)      # To do data science
library(tidylog)        # To provide feedback on dplyr functions
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(readxl)         # To read Excel files
library(digest)         # To generate hashes
library(rgbif)          # To use GBIF services
```

# Read and tidy source data

The dataset is managed in a Google Sheet that is [published to the web](https://docs.google.com/spreadsheets/d/e/2PACX-1vQTOIBGFbyhkhpLLzwryEvJ2NcR5C4-__iQwzw6NncQTQXSwDuHPfL_t43OMUTmEYBY7HnhaidG4K0J/pub?gid=818849950&single=true&output=csv) as a .csv.

```{r}
raw_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vQTOIBGFbyhkhpLLzwryEvJ2NcR5C4-__iQwzw6NncQTQXSwDuHPfL_t43OMUTmEYBY7HnhaidG4K0J/pub?gid=818849950&single=true&output=csv", col_types = cols(.default = "c"))
```

Preview data:

```{r}
raw_data %>% head(n = 5)
```

Clean data somewhat:

```{r}
input_data <-
  raw_data |>
  remove_empty("rows") |>
  clean_names() 
```

# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows). Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

Start with record-level terms which contain metadata about the dataset (which is generally the same for all records). Make sure to replace the dummy values with your own values.
Then the taxon-level terms will be added in the next step.

```{r}
taxon <-
  input_data |>
  distinct(taxon_id, .keep_all = TRUE) |>
  mutate(
    # RECORD-LEVEL
    language = "en",
    license = "CC BY 4.0",
    rightsHolder = "Ovidius University of Constanta",
    datasetID = "my_dataset_doi",
    institutionCode = "UoC",
    datasetName = "Checklist of introduced taxa in Romania", 
    # TAXON-LEVEL
    taxonID = taxon_id,
    scientificName = scientific_name,
    kingdom = kingdom,
    phylum = phylum,
    class = class,
    order = order,
    family = family,
    genus = genus,
    taxonRank = taxon_rank,
    nomenclaturalCode = nomenclatural_code
  ) |>
  # Arrange by taxonID and only keep the Darwin Core columns
  arrange(taxonID) |> 
  select(
    "language", "license", "rightsHolder", "datasetID", "institutionCode",
    "datasetName", "taxonID", "scientificName", "kingdom", "phylum", "class",
    "order", "family", "genus",
    "taxonRank", "nomenclaturalCode"
  )
```


## Distribution extension

Create a dataframe with all data (including multiple distributions). Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

```{r}
distribution <-
  input_data |>
  # pathway
  pivot_longer(
    names_to = "key", 
    values_to = "pathway", 
    starts_with("introduction_pathway"),
    values_drop_na = FALSE) |>
  filter( # keep NA value for species with no pathway provided
    !is.na(pathway) |
      (is.na(pathway) & key == "introduction_pathway_1")
      ) |> 
  # other terms
  mutate(
    taxonID = taxon_id,
    locality = location,
    countryCode = country_code,
    occurrenceStatus = occurrence_status,
    establishmentMeans = establishment_means,
    degreeOfEstablishment = degree_of_establishment,
    eventDate = case_when(
      is.na(date_first_observation) & is.na(date_last_observation) ~ NA,
      is.na(date_first_observation) ~ paste0("/", date_last_observation),
      is.na(date_last_observation) ~ paste0(date_first_observation, "/"),
      !is.na(date_first_observation) & !is.na(date_last_observation) ~ 
        paste(date_first_observation, date_last_observation, sep = "/")
    ),
    source = source,
    occurrenceRemarks = remarks
  ) |>
  # Arrange by taxonID and only keep the Darwin Core columns
  arrange(taxonID) |>
  select(
    "taxonID", "locality", "countryCode", "occurrenceStatus",
    "establishmentMeans", "degreeOfEstablishment", "pathway", "eventDate",
    "source", "occurrenceRemarks"
  ) 
```

## Species profile extension

In this extension we will express broad habitat characteristics of the species (e.g. `isTerrestrial`).

Create a dataframe with unique taxa only (ignoring multiple distribution rows).
Only keep records for which `terrestrial`, `marine` and `freshwater` is not empty.

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

```{r}
species_profile <-
  input_data |>
  distinct(taxon_id, .keep_all = TRUE) |> 
  filter(
    !is.na(terrestrial) |
      !is.na(marine) |
      !is.na(freshwater)
  ) |> 
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    isMarine = marine,
    isFreshwater = freshwater,
    isTerrestrial = terrestrial
  ) |> 
  arrange(taxonID)
```

## Description extension

In the description extension we want to include the native range of a species.

```{r}
description <-
  input_data |> 
  # unique taxa only (ignoring multiple distribution rows)
  distinct(taxon_id, .keep_all = TRUE) |> 
  # Separate values on `|` 
  mutate(native_range = strsplit(native_range, "\\|")) |> 
  unnest(native_range) |> 
  filter(!is.na(native_range)) |> 
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    description = str_trim(native_range),
    type = "native range",
    language = "en"
    ) |> 
  arrange(taxonID)
```

# Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description, here("data", "processed", "description.csv"), na = "")
```
